Miniproject - Analytic Number Theory

### 2021-12-18 Analytic number theory/submission.tex
Final corrections

    Now consider the element $\alpha = \omega_q - xqP \in \mathcal{O}$. Then
    \begin{align*}
        N_{K/\Q}(\alpha) &= \prod_{\sigma: K \to \C \ \text{$\Q$-field homomorphism}} \sigma(\omega_q - xqP) \\
        &= \prod_{\sigma} (\sigma(\omega_q) - xqP) = \mathrm{MiPo}_\Q(\omega_q)(xqP) = \Phi_q(xqP) \neq 1
    \end{align*}
    as $\mathrm{MiPo}_\Q(\omega_q) = \prod_\sigma (\sigma(\omega_q) - X)$.
    Hence, $\alpha$ is not a unit in $\mathcal{O}$.
    On the other hand, $(\alpha)$ is coprime to $(p_i)$ for each $p_i$, as 
    \begin{equation*}
        \omega_q = \alpha - xqP \in (\alpha) + (p_i) \quad \text{and} \quad \omega_q \in \mathcal{O}^\units
    \end{equation*}
    By our assumption, the only prime ideals in $\mathcal{O}$ are the prime ideal factors of $(p_i)$ and $(p)$ for $p \neq p_i$.
    Thus, the prime ideal factorization of $(\alpha)$ consists only of prime ideals $(p), p \neq p_i$ and it follows that $(\alpha) = (n)$ for some integer $n \geq 2$.
    As $\omega_q$ and $xqP \in \Z$ are $\Q$-linearly independent, we see that $n \divides \omega_q$ and $n \divides xqP$.
    However, the former is a contradiction, as $\omega_q \in \mathcal{O}^\units$ is a unit and no $n \geq 2$ is a unit.
\end{proof}
In a later chapter, the book also proves the general result \cite[Chapter VII, Theorem 5.14]{neukirch}.
I am pretty sure that those are exactly the functions $L(\chi, s)$ that we defined above.
\begin{corollary}[Question (iii)]
    There are infinitely many primes $p$ with $p \equiv 1 \pmod 4$.
\end{corollary}
\begin{proof}
    This is just a special case of Prop.~\ref{prop:primes_1_mod_q}.
\end{proof}
\begin{example}[Question (iii)]
    Using a computer, we can also study the actual frequency of prime numbers $\equiv 1, 3 \pmod 4$ among e.g. the first $10^8$ integers.
    This seems to indicate that both numbers are asymptotically equal, which seems natural, given the result of Example~\ref{ex:dirichlet_sum_over_primes}.
    For example, there are 332180 primes $\equiv 1 \pmod 4$ and 332398 primes $\equiv 3$ smaller than $10^8$.
    To find these numbers, the following python code was used.
    \lstinputlisting[language = python]{./primes.py}
\end{example}

@@

\section{Part III}
Again, let $q \geq 3$ be a prime.
Let further $\chi$ be a Dirichlet character mod $q$.
\begin{proposition}[Question (i)]
    \label{prop:logarithmic_derivative_L}
    For $\Re(s) > 1$ have that
    \begin{equation*}
        \frac {L(s, \chi)'} {L(s, \chi)} = \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s}
    \end{equation*}
\end{proposition}
\begin{proof}
    Consider any $\epsilon > 0$. The series
    \begin{equation*}
        \sum_{n \geq 1} \frac d {ds} \chi(n) n^{-s} = \sum_{n \geq 1} \chi(n) \log(n) n^{-s}
    \end{equation*}
    converges uniformly on $\Re(s) \geq 1 + \epsilon$, as $|\chi(n)| \leq C$ for some $C > 0$ and all $n \in \N$ (by the lecture, we know that $\sum_n \log(n) n^{-s}$ converges uniformly on $\Re(s) \geq 1 + \epsilon$).
    Hence, we may interchange summation and differentiation to get
    \begin{equation*}
        L(s, \chi)' = \sum_{n \geq 1} \chi(n) \frac d {ds} n^{-s} = \sum_{n \geq 1} \chi(n) \log(n) n^{-s}
    \end{equation*}
    for $\Re(s) \geq 1 + \epsilon$.
    As $\epsilon > 0$ was arbitrary, we get
    \begin{equation*}
        L(s, \chi)' = \sum_{n \geq 1} \chi(n) \log(n) n^{-s}
    \end{equation*}
    for all $\Re(s) > 1$.

@@

    However, the derivative of a holomorphic function is again holomorphic, so $L(s, \chi)'$ has no pole at $s = 1$.
    Provided that $L(1, \chi) \neq 0$, it follows that $L(s, \chi)'/L(s, \chi)$ is holomorphic on a compact neighborhood of $1$, so bounded.
\end{proof}
Now we can show the main result of this miniproject.
We will prove two auxiliary lemmas before.
\begin{lemma}
    \label{prop:rho_pole_at_one}
    For $a \perp q$, the function
    \begin{equation*}
        \rho_a(s) := \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s}
    \end{equation*}
    defined on $\Re(s) > 1$ has a meromorphic continuation to $\Re(s) > 0$ with a simple poles at $1$ (and possibly other poles on $\Re(s) > 0$), provided that $L(\chi, 1) \neq 0$ for all $\chi \in \mathrm{Dir}(q) \setminus \{ \chi_0 \}$.
\end{lemma}
\begin{proof}
    By Prop.~\ref{prop:logarithmic_derivative_L}, we have for $\chi \in \mathrm{Dir}(q)$ and $\Re(s) > 1$ that
    \begin{equation*}
        \frac {\overline{\chi(a)}} {q - 1} \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s} = \frac {\overline{\chi(a)}} {q - 1} \frac {L(s, \chi)'} {L(s, \chi)}
    \end{equation*}
    Hence
    \begin{equation*}
        \rho_a(s) := \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \frac {L(s, \chi)'} {L(s, \chi)}
    \end{equation*}
    for $\Re(s) > 1$ and the right-hand side is a meromorphic continuation of $\rho_a(s)$ to $\Re(s) > 0$.

@@

    Together, this yields that the sum of those functions
    \begin{equation*}
        \rho_a(s) = \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \frac {L(\chi, s)'} {L(\chi, s)}
    \end{equation*}
    is a meromorphic function with a simple pole at $s = 1$.
\end{proof}
\begin{lemma}
    \label{prop:correlation_theta_Psi}
    Let $a \perp q$ and define
    \begin{equation*}
        \Psi_a(x) := \sum_{n < x, \ n \equiv a \ (\text{mod} \ q)} \Lambda(n)
    \end{equation*}
    and
    \begin{equation*}
        \theta_a(x) := \sum_{p < x, \ p \equiv a \ (\text{mod} \ q)} \log(p)
    \end{equation*}
    Then
    \begin{equation*}
        \Psi_a(x) - \theta_a(x) = O(x^{1/2}\log(x))
    \end{equation*}
\end{lemma}
\begin{proof}
    Have
    \begin{align*}
        \Psi_a(x) - \theta_a(x) =& \sum_{p^k < x, \ p^k \equiv a \ (\text{mod} \ q)} \log(p) - \sum_{p < x, \ p \equiv a \ (\text{mod} \ q)} \log(p) \\
        =& \sum_{p^k < x, \ k \geq 2, \ p^k \equiv a \ (\text{mod} \ q)} \log(p) \leq \sum_{p^k < x, \ k \geq 2} \log(p) \\
        =& \Psi(x) - \theta(x) = O(x^{1/2}\log(x))
    \end{align*}
    where the last equality was proven in the lecture.
\end{proof}
\begin{proposition}[Question (v)]
    \label{prop:infinite_primes_congruent}
    Assume that $L(1, \chi) \neq 0$ for all $\chi \in \mathrm{Dir}(q) \setminus \{ \chi_0 \}$.
    Then for all $a \perp q$ there are infinitely many primes $p \equiv a \pmod q$.
\end{proposition}
\begin{proof}
    Assume not, then $\theta_a(x)$ is bounded, i.e. $\theta_a(x) = O(1)$.
    With Lemma~\ref{prop:correlation_theta_Psi} it follows that $\Psi_a(x) = O(x^{1/2} \log x)$.

>>>>

    Now consider the element $\alpha = \omega_q - xqP \in \mathcal{O}$. Then
    \begin{align*}
        N_{K/\Q}(\alpha) &= \prod_{\sigma: K \to \C \ \text{$\Q$-field homomorphism}} \sigma(\omega_q - xqP) \\
        &= \prod_{\sigma} (\sigma(\omega_q) - xqP) = \mathrm{MiPo}_\Q(\omega_q)(xqP) = \Phi_q(xqP) \neq 1
    \end{align*}
    as $\mathrm{MiPo}_\Q(\omega_q) = \prod_\sigma (\sigma(\omega_q) - X)$.
    Hence, $\alpha$ is not a unit in $\mathcal{O}$.
    On the other hand, $(\alpha)$ is coprime to $(p_i)$ for each $p_i$, as 
    \begin{equation*}
        \omega_q = \alpha - xqP \in (\alpha) + (p_i) \quad \text{and} \quad \omega_q \in \mathcal{O}^\units
    \end{equation*}
    By our assumption, the only prime ideals in $\mathcal{O}$ are the prime ideal factors of $(p_i)$ and $(p)$ for $p \neq p_i$.
    Thus, the prime ideal factorization of $(\alpha)$ consists only of prime ideals $(p), p \neq p_i$ and it follows that $(\alpha) = (n)$ for some integer $n \geq 2$.
    As $\omega_q$ and $xqP \in \Z$ are $\Q$-linearly independent, we see that $n \divides \omega_q$ and $n \divides xqP$.
    However, the former is a contradiction, as $\omega_q \in \mathcal{O}^\units$ is a unit and no $n \geq 2$ is a unit.
\end{proof}
In a later chapter, the book also proves the general result \cite[Chapter VII, Theorem 5.14]{neukirch}.
I am pretty sure that those are exactly the functions $L(s, \chi)$ that we defined above.
\begin{corollary}[Question (iii)]
    There are infinitely many primes $p$ with $p \equiv 1 \pmod 4$.
\end{corollary}
\begin{proof}
    This is just a special case of Prop.~\ref{prop:primes_1_mod_q}.
\end{proof}
\begin{example}[Question (iii)]
    Using a computer, we can also study the actual frequency of prime numbers $\equiv 1, 3 \pmod 4$ among e.g. the first $10^8$ integers.
    This seems to indicate that both numbers are asymptotically equal, which seems natural, given the result of Example~\ref{ex:dirichlet_sum_over_primes}.
    For example, there are 332180 primes $\equiv 1 \pmod 4$ and 332398 primes $\equiv 3$ smaller than $10^8$.
    To find these numbers, the following python code was used.
    \lstinputlisting[language = python]{./primes.py}
\end{example}

@@

\section{Part III}
Again, let $q \geq 3$ be a prime.
Let further $\chi$ be a Dirichlet character mod $q$.
\begin{proposition}[Question (i)]
    \label{prop:logarithmic_derivative_L}
    For $\Re(s) > 1$ have that
    \begin{equation*}
        \frac {L(s, \chi)'} {L(s, \chi)} = \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s}
    \end{equation*}
    Since the right-hand side converges absolutely for $\Re(s) > 1$, we see that $L(s, \chi)'/L(s, \chi)$ has no pole and so $L(s, \chi)$ is nonzero on $\Re(s) > 1$.
\end{proposition}
\begin{proof}
    Consider any $\epsilon > 0$. The series
    \begin{equation*}
        \sum_{n \geq 1} \frac d {ds} \chi(n) n^{-s} = \sum_{n \geq 1} \chi(n) \log(n) n^{-s}
    \end{equation*}
    converges uniformly on $\Re(s) \geq 1 + \epsilon$, as $|\chi(n)| \leq C$ for some $C > 0$ and all $n \in \N$ (by the lecture, we know that $\sum_n \log(n) n^{-s}$ converges uniformly on $\Re(s) \geq 1 + \epsilon$).
    Hence, we may interchange summation and differentiation to get
    \begin{equation*}
        L(s, \chi)' = \sum_{n \geq 1} \chi(n) \frac d {ds} n^{-s} = \sum_{n \geq 1} \chi(n) \log(n) n^{-s}
    \end{equation*}
    for $\Re(s) \geq 1 + \epsilon$.
    As $\epsilon > 0$ was arbitrary, we get
    \begin{equation*}
        L(s, \chi)' = \sum_{n \geq 1} \chi(n) \log(n) n^{-s}
    \end{equation*}
    for all $\Re(s) > 1$.

@@

    However, the derivative of a holomorphic function is again holomorphic, so $L(s, \chi)'$ has no pole at $s = 1$.
    Provided that $L(1, \chi) \neq 0$, it follows that $L(s, \chi)'/L(s, \chi)$ is holomorphic on a compact neighborhood of $1$, so bounded.
\end{proof}
Now we can show the main result of this miniproject.
We will prove two auxiliary lemmas before.
\begin{lemma}
    \label{prop:rho_pole_at_one}
    For $a \perp q$, the function
    \begin{equation*}
        \rho_a(s) := \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s}
    \end{equation*}
    defined on $\Re(s) > 1$ has a meromorphic continuation to $\Re(s) > 0$ with a simple poles at $1$ (and possibly other poles on $\Re(s) > 0$), provided that $L(1, \chi) \neq 0$ for all $\chi \in \mathrm{Dir}(q) \setminus \{ \chi_0 \}$.
\end{lemma}
\begin{proof}
    By Prop.~\ref{prop:logarithmic_derivative_L}, we have for $\chi \in \mathrm{Dir}(q)$ and $\Re(s) > 1$ that
    \begin{equation*}
        \frac {\overline{\chi(a)}} {q - 1} \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s} = \frac {\overline{\chi(a)}} {q - 1} \frac {L(s, \chi)'} {L(s, \chi)}
    \end{equation*}
    Hence
    \begin{equation*}
        \rho_a(s) := \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \frac {L(s, \chi)'} {L(s, \chi)}
    \end{equation*}
    for $\Re(s) > 1$ and the right-hand side is a meromorphic continuation of $\rho_a(s)$ to $\Re(s) > 0$.

@@

    Together, this yields that the sum of those functions
    \begin{equation*}
        \rho_a(s) = \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \frac {L(s, \chi)'} {L(s, \chi)}
    \end{equation*}
    is a meromorphic function with a simple pole at $s = 1$.
\end{proof}
\begin{lemma}
    \label{prop:correlation_theta_Psi}
    Let $a \perp q$ and define
    \begin{equation*}
        \Psi_a(x) := \sum_{n < x, \ n \equiv a \ (\text{mod} \ q)} \Lambda(n)
    \end{equation*}
    and
    \begin{equation*}
        \theta_a(x) := \sum_{p < x, \ p \equiv a \ (\text{mod} \ q)} \log(p)
    \end{equation*}
    Then
    \begin{equation*}
        \Psi_a(x) - \theta_a(x) = O(x^{1/2}\log(x))
    \end{equation*}
\end{lemma}
\begin{proof}
    Have
    \begin{align*}
        \Psi_a(x) - \theta_a(x) =& \sum_{p^k < x, \ p^k \equiv a \ (\text{mod} \ q)} \log(p) - \sum_{p < x, \ p \equiv a \ (\text{mod} \ q)} \log(p) \\
        =& \sum_{p^k < x, \ k \geq 2, \ p^k \equiv a \ (\text{mod} \ q)} \log(p) \leq \sum_{p^k < x, \ k \geq 2} \log(p) \\
        =& \Psi(x) - \theta(x) = O(x^{1/2}\log(x))
    \end{align*}
    where the last equality was proven in the lecture.
\end{proof}
\begin{proposition}[Question (v)]
    \label{prop:infinite_primes_congruent}
    Assume that $L(1, \chi) \neq 0$ for all $\chi \in \mathrm{Dir}(q) \setminus \{ \chi_0 \}$.
    Then for all $a \perp q$ there are infinitely many primes $p \equiv a \pmod q$.
\end{proposition}
\begin{proof}
    Assume not, then $\theta_a(x)$ is bounded, i.e. $\theta_a(x) = O(1)$.
    With Lemma~\ref{prop:correlation_theta_Psi} it follows that $\Psi_a(x) = O(x^{1/2} \log x)$.
